---
title: "Zinc in Elderly Project"
author: "Marco A Nunes"
date: "2 de novembro de 2018"
output: html_document
---
Linkedin: https://www.linkedin.com/in/marco-nunes-538a8b57/ <br>
GitHub: https://github.com/nunesma <br> <br>

### Zinc blood concentration in Elderly 

Objective: the aim of this research was to evaluate the blood concentration of zinc in elderly patients hospitalized in a general hospital

List of variables: <br>
- pac - pacient - patient record in the survey <br>
- sex - sex - registered as male or female <br>
- age - age - registered in years <br>
- icu - intensive care unit - hospitalization in an intensive care unit; registered as yes or no <br>
- ren - renal dysfunction - diagnosis of renal failure; registered as yes or no<br>
- alb - albumin - serum albumin concentration <br>
- lbd - low blood count - red blood cell count (yes: <11, no: >11) <br>
- hlk - high leukogram - leukocyte count (yes: >10.000, no: <10.000) <br>
- znc - zinc - serum zinc concentration <br>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
# input data
zinc <- read.csv("zinc.csv", sep =";")
```

Summary quantitative analysis of all variables showing an sick elderly sample.
```{r, echo=FALSE}
summary(zinc[-1])
```

Plot Series

```{r, echo=FALSE}
par(mfrow=c(1,3))
hist(zinc$age, main = "", xlab = "Age")
hist(zinc$alb, main = "", xlab = "Albumin")
hist(zinc$znc, main = "", xlab = "Zinc")
```


```{r, echo=FALSE}
par(mfrow=c(1,2))
plot(zinc$age,zinc$znc, main = "", ylab = "Zinc", xlab = "Age")
plot(zinc$alb, zinc$znc,main = "", ylab = "Zinc", xlab = "Albumin")
```

```{r, echo=FALSE}
par(mfrow=c(1,3))
boxplot(zinc$znc ~ zinc$sex, ylab = "Zinc", xlab = "Sex")
boxplot(zinc$znc ~ zinc$icu, ylab = "Zinc", xlab = "Intensive Care Unit")
boxplot(zinc$znc ~ zinc$ren, ylab = "Zinc", xlab = "Renal Dysfunction")
par(mfrow=c(1,2))
boxplot(zinc$znc ~ zinc$lbd, ylab = "Zinc", xlab = " Low Red Blood Cell Count")
boxplot(zinc$znc ~ zinc$hlk, ylab = "Zinc", xlab = "High Leukocyte Count")
```

Zinc concentration bivariate analysis
```{r}
# bivariate analysis
tapply(zinc$znc, zinc$sex, summary)
tapply(zinc$znc, zinc$icu, summary)
tapply(zinc$znc, zinc$ren, summary)
tapply(zinc$znc, zinc$lbd, summary)
tapply(zinc$znc, zinc$hlk, summary)
```

Related inference analysis
```{r, echo=FALSE}
# inference analysis
wilcox.test(zinc$znc ~ zinc$sex)
wilcox.test(zinc$znc ~ zinc$icu)
wilcox.test(zinc$znc ~ zinc$ren)
wilcox.test(zinc$znc ~ zinc$lbd)
wilcox.test(zinc$znc ~ zinc$hlk)
```


```{r, echo=FALSE}
# preparing dummy variables
zinc$sex <- as.numeric(as.character(zinc$sex) == "female")
zinc$icu <- as.numeric(as.character(zinc$icu) == "yes")
zinc$ren <- as.numeric(as.character(zinc$ren) == "yes")
zinc$lbd <- as.numeric(as.character(zinc$lbd) == "yes")
zinc$hlk <- as.numeric(as.character(zinc$hlk) == "yes")
```

Correlation matrix showing correlation absence.
```{r, echo=FALSE}
# correlation matrix
cor(zinc[,2:9])
```

Simple linear models
```{r}
# simple linear models
mod1 <- lm(znc ~ age, data = zinc); summary(mod1)
mod2 <- lm(znc ~ alb, data = zinc); summary(mod2)
mod3 <- lm(znc ~ sex, data = zinc); summary(mod3)
mod4 <- lm(znc ~ icu, data = zinc); summary(mod4)
mod5 <- lm(znc ~ ren, data = zinc); summary(mod5)
mod6 <- lm(znc ~ lbd, data = zinc); summary(mod6)
mod7 <- lm(znc ~ hlk, data = zinc); summary(mod7)
```

Multiple linear models
```{r}
# multiple linear models
mod8 <- lm(znc ~ age + alb + sex + icu + ren + lbd + hlk, data = zinc); summary(mod8)
```

### Conclusion:
These variables were not related with zinc concentration in elderly patients hospitalized in a general hospital
